BOLD Parameter Estimation Using Particle Filters

### 2010-07-10 doc/thesis/methods.tex
added quite a few plots. Should be able to write results for single voxel simulations.

\section{Prior Distribution}
For the BOLD model described in \autoref{sec:BOLD Physiology}, several
different studies have endeavored to calculate parameters. The results
of these studies may be found in \autoref{tab:Params}, and the methods 
used for each may be found in \autoref{sec:Prior Works}. Unfortunately,
\cite{Friston2000} only studied small regions; and most research with
reasonable speed (including \cite{Friston2002b}) used these results as 
the source for their priors. 
The one exception is \cite{Johnston2008}, which came to an extremely different
set of distributions. For a particle filter, the choice of a prior is
the single most important design choice. Consequently for this work,
I wanted to be conservative; which meant going with the accepted result,
\cite{Friston2000}. This constrains the usefulness of the model to
areas that fall within the prior distribution, yet will allow results
to be comparable to other works. There is a significant need for better
estimates of the physiological parameters; and, while physical experiments
may not be possible, it would not be unreasonable to do a study with
exhaustive simulated annealing or hill climbing tests for multiple
regions and multiple patients. The purpose for this work is to determine
the fitness of particle filters for this task; because with good 
priors it can be extremely fast. Ultimately this algorithm may become
more useful as more studies are done on BOLD model that don't require
priors.

Therefore, because \cite{Friston2000} represents the current best
knowledge about the BOLD parameters, it forms the basis
for the prior distributions in this paper.

\begin{table}[t]
\centering
\begin{tabular}{|c || c | c | c |}
\hline 
Parameter & Distribution & $\mu$ & $\sigma$ \\
\hline
$\tau_0$ & Gamma & .98 & .5  \\
$\alpha$ & Gamma & .33 & .045\\
$E_0$    & Gamma & .34 & .03  \\
$V_0$    & Gamma & .03 & .006 \\
$\tau_s$ & Gamma & 1.54  & .5 \\
$\tau_f$ & Gamma & 2.46  & .5 \\
$\epsilon$ & Gamma & .64 & .4 \\
\hline
\end{tabular}
\caption{}
\label{tab:Prior} 
\end{table}

>>>>

\section{Prior Distribution}
\label{sec:PriorDist}
For the BOLD model described in \autoref{sec:BOLD Physiology}, several
different studies have endeavored to calculate parameters. The results
of these studies may be found in \autoref{tab:Params}, and the methods 
used for each may be found in \autoref{sec:Prior Works}. Unfortunately,
\cite{Friston2000} only studied small regions; and most research with
reasonable speed (including \cite{Friston2002b}) used these results as 
the source for their priors. 
The one exception is \cite{Johnston2008}, which came to an extremely different
distributions. For a particle filter, the choice of a prior is
the single most important design choice. Consequently for this work,
I wanted to be conservative; which meant going with the accepted result,
\cite{Friston2000}. This constrains the usefulness of the model to
areas that fall within the prior distribution, yet will allow results
to be comparable to other works. There is a significant need for better
estimates of the physiological parameters; and, while physical experiments
may not be possible, it would not be unreasonable to do a study with
exhaustive simulated annealing or hill climbing tests for multiple
regions and multiple patients. The purpose for this work is to determine
the fitness of particle filters for this task; because with good 
priors it can be extremely fast. Ultimately this algorithm may become
more useful as more studies are done on BOLD model that don't require
priors.

There is an interesting anomaly with the priors found in virtually all
the works that characterized the parameters, except \cite{Johnston2008}.
The BOLD signal is universally recognized to be around $2-3\%$, maybe
reaching $5\%$ in extreme activation. Yet using the mean priors
from \cite{Friston2000}, the signal response for a $.1$ second
impulse only reaches maybe half a percent, as \autoref{fig:MeanResponse}
shows.

\begin{figure}
\centering
\label{fig:MeanResponse}
\includegraphics[trim=6cm 3cm 6cm 3cm,width=16cm]{images/mean_response}
\caption{Response to $.1s$ impulses with the mean parameters from \cite{Friston2000}}
\end{figure}

While this could be the result of a stimulus
being too short to lead to strong activation, a similar stimulus
scheme in real data showed a significantly larger response than 
half a percent as well. In fact, after applying de-trending,
converting the image to percent-difference, and removing 
outliers ($ BOLD > 10\% \text{ or } BOLD < -10\%$) the total variance
across all \emph{active} voxels was still around .02, indicating
that in active voxels a signal peaking below .005 seems unlikely. 
Of course, if more restriction were placed on the outliers, its possible
this standard deviations could be brought down. 
The parameter estimates by \cite{Johnston2008} are even more 
confusing, with peaks of well below $.1\%$ (\autoref{fig:MeanResponseJ}).

\begin{figure}
\centering
\label{fig:MeanResponse}
\includegraphics[trim=6cm 3cm 6cm 3cm,width=16cm]{images/mean_response_johnston}
\caption{Response to $.1s$ impulses with the mean parameters from \cite{Johnston2008}}
\end{figure}

Its likely that these differences are due to some difference in preprocessing,
although in \cite{Deneux2006} the signals were found to be peaking around
$1\%$, unlike \cite{Friston2000} which shows signals peaking at up to
$3\%$ or $4\%$. In my own tests, it seemed necessary for $\epsilon$ to
reach well over $1.5$ and $V_0$ to reach more than $.4$ to reach these
peaks; of course other methods may be equally able. 
Therefore, to account for these discrepancies, somewhat broader
distributions are used than the numbers used in \cite{Friston2000}
(which are widely used, \cite{Hu2009}). The 
priors used in the particle filter may be found in \autoref{tab:Prior}.

\begin{table}[t]
\centering
\begin{tabular}{|c || c | c | c |}
\hline 
Parameter & Distribution & $\mu$ & $\sigma$ \\
\hline
$\tau_0$ & Gamma & .98 & .25 \\
$\alpha$ & Gamma & .33 & .045\\
$E_0$    & Gamma & .34 & .03  \\
$V_0$    & Gamma & .04 & .03 \\
$\tau_s$ & Gamma & 1.54  & .25\\
$\tau_f$ & Gamma & 2.46  & .25\\
$\epsilon$ & Gamma & .7  & .6 \\
\hline
\end{tabular}
\caption{}
\label{tab:Prior} 
\end{table}

### 2010-07-10 doc/thesis/results.tex
added quite a few plots. Should be able to write results for single voxel simulations.

\section{Simulation}
I performed two types of simulations. First, I simulated a single BOLD time-series based
on a randomly selected set of model parameters. This process was relatively straight forward,
given the state-space equations for the BOLD signal. After a "true" signal was generated,
I then added a carrier level, since BOLD is typically measured as a \% difference from the
base level. I tested the particle filter with two types of noise; independent and identically
distributed gaussian noise and weiner drift. For both I tested with various standard deviations.

Once this noisy simulated time series was generated, the exact same particle filter algorithm
that would later be run on full sized images, was run on this single voxel image. I ran
a series of tests to determine the convergence rate of the particle filter, the number
of particles that were required, how weighting functions compared, how different de-trending
methods compared with each other and, finally the variance of the result. By running the exact
time-series with different noise realizations, it was possible to determine the model variance.
As the reader may know, the error of an estimator may be calculated as:
\begin{equation}
MSE(\Theta) = Var(\Theta) + Bias(\Theta)^2
\end{equation}
The variance is an expression of how much the result would change for different noise realizations,
whereas the bias is an expression of how well the model matches the true underlying model. In
this case, because the same model is being used in the particle filter and underlying simulation,
the bias is actually zero. Obviously when this is calculated using \emph{real} data with an unknown
underlying state space equation, there will be some amount of bias error, but assuming that the
noise is similar to the noise used in these tests, the model variance will actually be about the
same. Thus calculating the model variance is extremely helpful in calculating how well determined
our model is, and how consistent it will be for real data. A single timeseries, as opposed to the
thousands present in a real image, makes it easier to 
compare the output with the ground truth, with various parameters.

@@

Another crucial factor to run time is how long before the first re-sampling 
occurs. Because the prior is represented initially with significantly more
particles, if the model fits very well, or for some reason the effective
number of particles stays high, resampling could take a long time to occur.
When this happens the particles filter can take a factor of 10 longer to run.
However, if the particle count isn't initially set high, there is a much larger
chance of particle deprivation occurring. Since there is no real way to know
how long it will take to resmaple the first time, there is little the 
algorithm can do to fix this. On the other end of the specture, if the time
series don't match the model at all, particle deprivation will occur extremely
quickly, and even the recovery technique discussed in \autoref{sec:Resampling}
won't help. The upshot of this is that the particle filter is able to 
identify these sections very quickly, and thus not waste much time there.
Of course, if a fat-tailed distribution is used for the weighting function,
or the standard deviation of a Gaussian weighting function is very large,
the particle filter will simply converge to meaningless values.

\begin{figure}
\caption{Particle Filter converging to values that make little sense,
because the voxel did not correlate with the input in any known way}
\end{figure}

>>>>

\section{Simulation}
I performed two types of simulations. First, I simulated a single BOLD time-series based
on a randomly selected set of model parameters. Second I used a modified version of the
FSL tool possum to generate a noisy FMRI image.

\subsection{Single Timeseries}
This process was relatively straight forward,
given the state-space equations for the BOLD signal. After a "true" signal was generated,
I then added a identically and independently distributed (I.I.D.) Gaussian noise and a Wiener
process with Gaussian I.I.D. steps. Finally a carrier level was added, since BOLD is typically
measured as a \% difference from the
base level. Adding a carrier level meant that the exact same algorithm which was used for 
real data could be used for the simulated data.

Once this noisy simulated time series was generated, the particle filter algorithm
 was run on this single timeseries image. In order to find the best setting for the
particle filter I ran quite a few tests. Here I will include two sets of tests 
to determine the power of the particle filter in modeling. The first test is
low noise $\{\sigma_y = .001, \sigma_x = .0005\}$, while the second set of
tests had noise of $ \{\sigma_y = .01, \sigma_x = .005\} $, where $\sigma_y$ is the
measurement noise, and $\sigma_x$ is the wiener step size. Both signals used the
parameters ($\tau_0 = 1.45, \alpha = .3, E_0 = .47, V_0 = .044, \tau_s = 1.94, \tau_f = 1.99, \epsilon = 1.8$).
The particle filter used the parameters defined in \autoref{tab:Prior} (\autoref{sec:PriorDistrib}),
thus the particle filter was not centered over the correct values.

\subsubsection{Low Noise Simulation}
For the low noise case, the ten realization are shown in \autoref{fig:LowNoiseRealization}.
\begin{figure}
\label{fig:LowNoiseRealization}
\includegraphics[trim=6cm 3cm 6cm 3cm,width=16cm]{images/realization_lownoise}
\caption{Test Signals with low noise compared to the clean signal}
\end{figure}

The resulting fits compared to the \emph{true} BOLD signal are shown in \autoref{fig:FitComparisonLowNoise}.
The fits are good, but the noise is relatively low. The preprocessed signal compared
to the true BOLD signal is shown in \autoref{fig:PreprocessedLowNoise}.

\begin{figure}
\label{fig:PreprocessedLowNoise}
\includegraphics[trim=6cm 3cm 6cm 3cm,width=16cm]{images/preprocessed_lownoise}
\caption{A comparison of the preprocessed signals for the low noise case.}
\end{figure}

The preprocessing consists of several steps, which are discussed in detail in \autoref{sec:Preprocessing}.
Obviously the spline struggles a little bit at the end, as the graph shows, although 
overall the fit is pretty well. Finally, the set of fits to the preprocessed data, and in 
essence to the "true" BOLD signal is shown in \autoref{fig:FitComparisonLowNoise}.

\begin{figure}
\label{fig:FitComparisonLowNoise}
\includegraphics[trim=6cm 3cm 6cm 3cm,width=16cm]{images/comparison_lownoise}
\caption{A comparison of the fitted signals for the low noise case.}
\end{figure}

It becomes clear now the reason for the term particle filter. In several locations the output
of the particle filter looks like a filtered version of the input. For instance toward the
end, the estimates stay flat in spite of the preprocessed data drifting off a bit. By
this point, the algorithm had converged sufficiently to know that that wasn't possible.
A similar circumstance occurs at 100 seconds in. In general the fits are good, although because
of the slightly higher signal levels in the preprocessed signals, the fits tend to overestimate
the peak a bit. The final parameter sets are shown in \autoref{tab:LowNoiseParams}
\begin{table}[t]
\centering
\begin{tabular}{|c | c | c | c | c | c | c | c | c | c |}
\hline 
$\tau_0$ & $\alpha$ & $E_0$    & $V_0$    & $\tau_s$ & $\tau_f$ & $\epsilon$  & $ \sum \tau $ & $\sqrt{MSE}$ (Res.) &$\sqrt{MSE}$\\
\hline 
\rowcolor[gray]{.8}
1.45 & .3 & .47 & .044 & 1.94 & 1.99 & 1.8  & 5.38 &  & \\
\hline 
\hline 
1.22214 & 0.3449 & 0.33462 & 0.07138 & 1.60446 & 2.27529 & 1.59454 & 5.10189  &  0.00321067  & 0.00987647  \\
1.37493 & 0.33183 & 0.36296 & 0.07327 & 1.64076 & 2.10299 & 1.57627 & 5.11867 &  0.00305475  & 0.00993227  \\
1.16604 & 0.32205 & 0.34057 & 0.08217 & 1.64768 & 2.35351 & 1.24515 & 5.16722 &  0.00328932  & 0.00967961  \\
1.23176 & 0.32707 & 0.34028 & 0.07961 & 1.62698 & 2.18519 & 1.30326 & 5.04394 &  0.00284719  & 0.00912005  \\
1.1832 & 0.31787 & 0.34718 & 0.0821 & 1.54961 & 2.29115 & 1.27817 & 5.02396   &  0.00300634  & 0.00971258  \\
1.1424 & 0.33395 & 0.34725 & 0.07366 & 1.62208 & 2.29084 & 1.4025 & 5.05531   &  0.00283287  & 0.00948483  \\
1.30041 & 0.3596 & 0.35643 & 0.07679 & 1.56406 & 2.13234 & 1.60338 & 4.99681  &  0.00302802  & 0.01021885  \\
1.24008 & 0.34601 & 0.33978 & 0.08903 & 1.64989 & 2.23655 & 1.29004 & 5.12651 &  0.00304378  & 0.01007964  \\
1.1709 & 0.32739 & 0.34644 & 0.08255 & 1.53734 & 2.28262 & 1.37828 & 4.99087  &  0.00334488  & 0.01032886  \\
1.18967 & 0.34344 & 0.33554 & 0.07976 & 1.53582 & 2.30746 & 1.42774 & 5.03295 &  0.00317542  & 0.01001503  \\
1.184 & 0.34053 & 0.35017 & 0.08917 & 1.61025 & 2.27926 & 1.16448 & 5.07352   &  0.00288855  & 0.00950536  \\
\hline                                                                           
1.21868 & 0.33588 & 0.34557 & 0.07995 & 1.59899 & 2.24884 & 1.38762 & 5.06651 & 0.00306562     & 0.00981396 \\
\hline 
\end{tabular}
\caption{Estimated Parameters on 10 different runs with low noise. First row is the true parameters,
last is mean over the 10 runs. The $\sqrt{MSE}$ (Res.) is the square root of the mean square of the
residuals. Essentially this is the is the MSE between the estimated signal and the noisy signal which 
was available to the particle filter. Square root of MSE is the actual \emph{error}, with the true signal,
which of course was not available to the particle filter.}
\label{tab:LowNoiseResults} 
\end{table}

There are a few things worth noting here. First the time constants vary greatly across
runs with different noise realizations, yet the sum of the individual time constants
, $\tau_f$, $\tau_s$ and $\tau_0$ seem to be more consistent. In general the 
time constants are falling short of the true time constant. This could be a limitation
based on the prior distribution (which notably has mean values below the values used
in the simulation) or it could be caused by the delayed benefit of having a 
correct time constant. Even if one particle has a better time constant than another, if
the difference isn't severe, by the time this makes a difference in the weight, the 
particle with the better $\tau$ will have been weighted based on other parameters several
times. Another interesting result in the huge variation in the levels of $V_0$. In general,
with this admittedly small amount of noise, it would appear that the relation between a set
of parameters/stimuli with the output is not injective. In other words, it would appear
that a timeseries is not unique to a set of parameters. This is good justification that 
simultaneous blood volume or tagged flow calculations with the conventional FMRI 
could definitely benefit the model.

\begin{table}[t]
\centering
\begin{tabular}{|c | c  c  c  c  c  c  c |}
\hline 
  & $\tau_0$ & $\alpha$ & $E_0$    & $V_0$    & $\tau_s$ & $\tau_f$ & $\epsilon$ \\
\hline 
\rowcolor[gray]{.8} $\tau_0$ &   0.0189481 & -0.0014269 & -0.0011267 & -1.13e-05 & -0.0025616 & -0.0189559 & 0.0070405 \\
$\alpha$ &                       -0.0014269 & 0.0026716 & 9.93e-05 & -0.0002041 & -0.0008632 & -0.0016823 & 0.0071891 \\
\rowcolor[gray]{.8} $E_0$    &   -0.0011267 & 9.93e-05 & 0.0010701 & -0.0002277 & -0.0001177 & 0.0001013 & 0.0016972 \\
$V_0$    &                       -1.13e-05 & -0.0002041 & -0.0002277 & 0.0005401 & 4.3e-06 & 4.56e-05 & -0.0080494 \\
\rowcolor[gray]{.8} $\tau_s$ &   -0.0025616 & -0.0008632 & -0.0001177 & 4.3e-06 & 0.0128056 & 0.012878 & -0.005516 \\
$\tau_f$ &                       -0.0189559 & -0.0016823 & 0.0001013 & 4.56e-05 & 0.012878 & 0.0416927 & -0.0158182 \\
\rowcolor[gray]{.8} $\epsilon$&  0.0070405 & 0.0071891 & 0.0016972 & -0.0080494 & -0.005516 & -0.0158182 & 0.1567165 \\
\hline 
\end{tabular}
\caption{Typical Covariance matrix of the parameters at the end of a run.}
\label{tab:CovSim} 
\end{table}

\begin{figure}[H]
\subfigure[Converging histogram for $\tau_0$, $\alpha$, $E_0$, and $V_0$ of the first run, low noise simulation.]
{\includegraphics[trim=7cm 4cm 7cm 4cm, width=15cm]{images/converge_lownoise1}}\\

\subfigure[Converging histogram for $\tau_s$, $\tau_f$, $\epsilon$, and $V$ of the first run, low noise simulation.]
{\includegraphics[trim=7cm 4cm 7cm 1cm, width=15cm]{images/converge_lownoise2}}\\
\end{figure}

\begin{figure}
\subfigure[Converging histogram for $Q$, $S$, $F$, and $BOLD$ of the first run, low noise simulation.]
{\includegraphics[trim=7cm 2cm 7cm 2cm, width=15cm]{images/converge_lownoise3}}
\label{fig:LowNoiseHist}
\end{figure}
The covariance matrix (\autoref{tab:CovSim}) also confirms the idea that the $\tau$ parameters are interchangeable
as far as the BOLD signal goes. Notice the covariance of $\tau_f$ and $\tau_0$ is $-0.019$ whereas
the variance of $\tau_0$ and $\tau_f$ are $0.019$ and $0.04$ respectively. Clearly they are
related. The convergence properties of the first run in \autoref{tab:LowNoiseResults} may be
viewed in \autoref{fig:LowNoiseHist}.

\subsubsection{High Noise Simulation}
For the high noise simulation, the exact same procedure was performed again, but with 
measurement error and drift standard
deviations an order of magnitude higher. The noisy signals versus the base signal are shown
in \autoref{fig:HighNoiseRealization} and the preprocessed signals which the particle
filter attempt to fit are in \autoref{fig:PreprocessedHighNoise}.
\begin{figure}[H]
\label{fig:LowNoiseRealization}
\includegraphics[trim=6cm 3cm 6cm 3cm,width=16cm]{images/realization_highnoise}
\caption{Test Signals with high noise compared to the clean signal, $\sigma_x = .01, \sigma_y=.005$}
\end{figure}
\begin{figure}[H]
\label{fig:PreprocessedHighNoise}
\includegraphics[trim=6cm 3cm 6cm 3cm,width=16cm]{images/preprocessed_highnoise}
\caption{A comparison of the preprocessed signals for the high noise case.}
\end{figure}

The results of the particle filter, for each of the ten runs may be seen in 
\autoref{fig:FitComparisonHighNoise}. Clearly the preprocessing led the algorithm
to somewhat higher activation levels, and it would appear that the subtleties of
different time constants are also lost for most of the runs, although a few seem
to manage quite accurate results.

\begin{figure}[H]
\label{fig:FitComparisonHighNoise}
\includegraphics[trim=6cm 3cm 6cm 3cm,width=16cm]{images/comparison_highnoise}
\caption{A comparison of the fitted signals for the high noise case.}
\end{figure}
\begin{figure}[H]
\label{fig:NoiseComparisonJustTwo}
\includegraphics[trim=6cm 3cm 6cm 3cm,width=16cm]{images/highnoise_56_noise}
\caption{Two particular preprocessed noise realizations for the high noise case.}
\end{figure}
\begin{figure}[H]
\label{fig:FitComparisonHighNoiseJust2}
\includegraphics[trim=6cm 3cm 6cm 3cm,width=16cm]{images/comparison_highnoise_just2}
\caption{The results for the noise realizations shown in \autoref{fig:NoiseComparisonJustTwo}.}
\end{figure}

It is interesting to consider how the preprocessing and noise may effect the
parameters of the fitting model. For instance run 1 in \autoref{fig:NoiseComparisonJustTwo}
certainly seems to be more biased toward higher peaks than run 2. There also appears
to be more drift than the 20 points per knot could fit, which explains the 
prolonged increase at 170 seconds. Despite run 1's bias toward higher peaks, for
some reason the particle filter was able to get a much better match for the
modeled post-stimulus undershoot (which is likely significantly shorter than it
should be). Ultimately the results are pretty good. \autoref{tab:SimMSE} shows
the mean squared error for all two runs, and highlights the two runs analyzed here and 
in \autoref{fig:ConvergenceRuns1} and \autoref{fig:ConvergenceRuns2}.

\begin{figure}[H]
\subfigure[$\tau_0$, $\alpha$, $E_0$, $V_0$]
{\includegraphics[trim=7cm 3cm 7cm 4cm, width=15cm]{images/highnoise_run5_1}}\\
\end{figure}
\begin{figure}[H]
\subfigure[$\tau_s$, $\tau_f$, $\epsilon$, $V$] 
{\includegraphics[trim=7cm 3cm 7cm 1cm, width=15cm]{images/highnoise_run5_2}}\\
\end{figure}
\begin{figure}[H]
\subfigure[$Q$, $S$, $F$, $BOLD$ ]
{\includegraphics[trim=7cm 3cm 7cm 2cm, width=15cm]{images/highnoise_run5_3}}
\label{fig:ConvergenceRuns1}
\caption{Converging histogram for parameters during run 2, as in \autoref{fig:NoiseComparisonJustTwo}.}
\end{figure}

There are a number of interesting convergence properties of the
particle filter when more noise is present, as both \autoref{fig:ConvergenceRuns1} and
\autoref{fig:ConvergenceRuns2} show. Obviously the particle filter seems to converge
significantly faster, as points tend to be further out on the weighting function. This
also causes significantly more resampling which is the explanation for the seeming
jumps in resolution that occur from time to time.

\begin{figure}[H]
\subfigure[$\tau_0$, $\alpha$, $E_0$, $V_0$]
{\includegraphics[trim=7cm 3cm 7cm 4cm, width=15cm]{images/highnoise_run6_1}}\\
\end{figure}
\begin{figure}[H]
\subfigure[$\tau_s$, $\tau_f$, $\epsilon$, $V$] 
{\includegraphics[trim=7cm 3cm 7cm 1cm, width=15cm]{images/highnoise_run6_2}}\\
\end{figure}
\begin{figure}[H]
\subfigure[$Q$, $S$, $F$, $BOLD$ ]
{\includegraphics[trim=7cm 3cm 7cm 2cm, width=15cm]{images/highnoise_run6_3}}
\label{fig:ConvergenceRuns2}
\caption{Converging histogram for parameters during run 2, as in \autoref{fig:NoiseComparisonJustTwo}.}
\end{figure}

The parameters arrived at for all ten filter runs are shown in \autoref{tab:HighNoiseResults}.

\begin{table}[t]
\centering
\begin{tabular}{|c | c | c | c | c | c | c | c | c | c |}
\hline 
$\tau_0$ & $\alpha$ & $E_0$    & $V_0$    & $\tau_s$ & $\tau_f$ & $\epsilon$  & $ \sum \tau $ & $\sqrt{MSE}$ (Res.) & $\sqrt{MSE}$\\
\hline 
\rowcolor[gray]{.8}
1.45 & .3 & .47 & .044 & 1.94 & 1.99 & 1.8  & 5.38 &  & \\
\hline 
\hline 
1.19002 & 0.23495 & 0.42228 & 0.128 & 1.01468 & 2.47795 & 1.11677 & 4.68265   &0.0140648  &0.01572919   \\
0.9721 & 0.21902 & 0.30505 & 0.061 & 0.57801 & 1.99596 & 3.46135 & 3.54608    &0.0137314  &0.01377685   \\
1.57947 & 0.14153 & 0.33798 & 0.1079 & 0.5843 & 2.12475 & 1.78343 & 4.28852   &0.01275372 &0.01577449   \\
1.10937 & 0.2374 & 0.53491 & 0.0351 & 1.21862 & 3.07365 & 2.35039 & 5.40164   &0.0167258  &0.0115433    \\
1.10712 & 0.27535 & 0.33651 & 0.03161 & 1.50567 & 2.65181 & 4.19099 & 5.2646  &0.01369793 &0.01221607   \\
0.58026 & 0.47931 & 0.41354 & 0.11894 & 0.97563 & 3.69018 & 1.00084 & 5.24607 &0.01149511 &0.01315637   \\
\rowcolor[rgb]{.9,.5,.5}
1.29515 & 0.25957 & 0.27559 & 0.25952 & 1.70265 & 2.8458 & 0.66172 & 5.84361  &0.01555035 &0.01789781   \\
\rowcolor[rgb]{.5,.5,.9}
1.5185 & 0.21987 & 0.28348 & 0.07417 & 0.88822 & 3.07706 & 1.73929 & 5.48378  &0.01205351 &0.01246202   \\
0.68736 & 0.3283 & 0.39786 & 0.15614 & 1.07781 & 3.1158 & 0.66432 & 4.88097   &0.01510364 &0.01257713   \\
1.01703 & 0.28497 & 0.34741 & 0.05672 & 1.58774 & 2.65157 & 2.28519 & 5.25634 &0.012493   &0.01343459   \\
0.99247 & 0.29795 & 0.32207 & 0.20943 & 0.42757 & 2.21081 & 1.01674 & 3.63085 &0.01216522 &0.01505545   \\
\hline                                                                          
1.09535 & 0.27075 & 0.36152 & 0.11259 & 1.05099 & 2.71958 & 1.84282 & 4.86592 &0.01362132  &0.01396575\\
\hline 
\end{tabular}
\caption{Estimated Parameters on 10 different runs with high noise. First row is the true parameters.
Note also that the blue row is Run 1 and the red row is Run 2, as used named this section}
\label{tab:HighNoiseResults} 
\end{table}

a series of tests to determine the convergence rate of the particle filter, the number
of particles that were required, how weighting functions compared, how different de-trending
methods compared with each other and, finally the variance of the result. By running the exact
time-series with different noise realizations, it was possible to determine the model variance.
As the reader may know, the error of an estimator may be calculated as:
\begin{equation}
MSE(\Theta) = Var(\Theta) + Bias(\Theta)^2
\end{equation}
The variance is an expression of how much the result would change for different noise realizations,
whereas the bias is an expression of how well the model matches the true underlying model. In
this case, because the same model is being used in the particle filter and underlying simulation,
the bias is actually zero. Obviously when this is calculated using \emph{real} data with an unknown
underlying state space equation, there will be some amount of bias error, but assuming that the
noise is similar to the noise used in these tests, the model variance will actually be about the
same. Thus calculating the model variance is extremely helpful in calculating how well determined
our model is, and how consistent it will be for real data. A single timeseries, as opposed to the
thousands present in a real image, makes it easier to 
compare the output with the ground truth, with various parameters.

@@

Another crucial factor for run time is how long before the first re-sampling 
occurs. Because the prior is represented initially with significantly more
particles, if the model fits very well, or for some reason the effective
number of particles stays high, resampling could take a long time to occur.
When this happens the particles filter can take a factor of 10 longer to run.
However, if the particle count isn't initially set high, there is a much larger
chance of particle deprivation occurring. Since there is no real way to know
how long it will take to resmaple the first time, there is little the 
algorithm can do to fix this (except perhaps forcing resampling after
some period of time). On the other end of the spectrum, if the time
series doesn't match the model at all, particle deprivation will occur extremely
quickly.  The upshot of this is that the particle filter is able to 
identify these sections very quickly, and thus not waste much time there.
The difficulty though, is the regions in between the perfect fit and the
awful fit. If particle deprivation does occur, did it occur randomly to an
activated region or did it occur inevitably because the region doesn't fit.
The reason for having so many initial particles is to give density to the
distribution to make false negatives more rare. Perhaps the correct method
is to still set the initial particles very high, but if for a long time no
resampling occurs, halve the standard deviation of the measurement error.
Thus, if the weighting function is not discriminating enough, force it 
to be more picky about the results. Of course this could lead to particle
deprivation as well if the standard deviation is brought down too quickly.
Of course, if a fat-tailed distribution is used for the weighting function,
or the standard deviation of a Gaussian weighting function is sufficiently large,
the particle filter will simply converge to meaningless values. The question
of whether

\begin{figure}
\label{fig:badfit}
\includegraphics[trim=10cm 4cm 10cm 4cm, width=16cm]{images/inactive_illogical}
\caption{Particle Filter converging to values that make little sense,
because the voxel did not correlate with the input in any known way}
\end{figure}

\begin{figure}
\label{fig:highhard}
\includegraphics[trim=10cm 4cm 10cm 4cm, width=16cm]{images/active_difficult}
\caption{Example of a timeseries that laplace performs better on than
gaussian. The large spike at the beginning can quickly eliminate all particles}

\end{figure}

For this reason, it is important to threshold based on the MSE, and the
impulse response. Naturally the MSE is intended to catch very voxels with a
very poor fit,
but like \autoref{fig:badfit}, the algorithm may have found a line
down the middle that gives a decent mean squared error. In this case
the excessive smoothing and low actual activation level will easily
be distinguished by finding the peak response to single short pulse.

The choice of a prior, as discussed previously, is extremely important. While a
prior may have the potential to give good results, being a monte-carlo algorithm
there is the possibility for inconsistencies. Thus, increasing the variance
of the time-constants may allow additional flexibility, it will also cause
additional model variance. Case in point, the exact same algorithm run twice
with standard deviations of $.35, .35, .35$ for the time constants resulted in two
very different fits, see \autoref{fig:param1_var}.

\begin{figure}
\label{fig:badfit_param1}
\subfigure[]{\includegraphics{images/badfit_param1}}
\subfigure[]{\includegraphics{images/goodfit_param1}}
\caption{The same priors gave rise to both fits. Params: ($\tau_0 \alpha E_0 V_0 \tau_f \tau_s \epsilon$)
for the good fit: $(1.40 .32 0.34 0.04  0.96 2.76)$, and for the poor fit:
$(2.28 0.32 0.33 0.017 0.67 2.88 1.36)$.}
\end{figure}

For this reason, I actually lowered the standard deviattions of the time
constants to prevent over-smoothing. This resulted in the more consistent, but
slightly worse results seen in \autoref{fig:param2_var}

\begin{figure}
\label{fig:param2_var}
\subfigure[]{\includegraphics{images/param2a}}
\subfigure[]{\includegraphics{images/param2b}}
\caption{A poor fit, using the same parameters as }
\end{figure}

\section{Particle Collapse Recovery}
This heuristic was initially intended to deal with cases where the particle
filter prematurely converged; and to reverse this effect. In reality, however
it did not perform well, and so it was not used in full-brain analysis. Ultimately
it did not prevent collapse of the distribution, and it often caused dramatic
increased in the variance of parameter estimates. It also pronged the analysis
time of inactive time series. Essentially the effect was exactly opposite of what
one would hope.
